The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2020-09-11 Writeup/manuscript.tex
Fixing title page
minor edits
wrapping figures.
edits to the Discussion ... close now

%\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission
\usepackage{graphicx}

@@

% \maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

@@

\begin{figure}[tbhp]
\centering
\includegraphics[width=1\linewidth]{../figures/figure1.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a P1 exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:durations_rotations}
\end{figure}

@@

We now wanted to quantify the degree to which our two measurements were consistent with the subgroup relationships. We hypothesized that more complex groups would (a) produce symmetric-specific SSVEPs with higher amplitudes and (b) shorter display duration thresholds. [PERHAPS WE NEED TO JUSTIFY THESE ASSUMPTIONS] We tested each of these hypotheses using the same approach. We first fitted a Bayesian model with wallpaper group as a factor and participant as a random effect. We fit the model separately for SSVEP RMS and psychophysical data, and then computed posterior distributions for the difference between supergroup and subgroup. These difference distributions could allowed us to compute the conditional probability that the supergroup would produce (a) larger RMS and (b) a shorter threshold durations, when compared to the subgroup. The posterior distributions are shown in Figure 2 for the SSVEP data, and in Figure 3 for the psychophysical data, which distributions color-coded according to conditional probability. For both data sets our hypothesis is confirmed: For the overwhelming majority of the 64 subgroup relationships, supergroups are more likely to produce larger symmetry specific SSVEPs and shorter threshold durations, and in most cases the conditional probability of this happening is extremely high.

@@

SSVEP data from four of the wallpaper groups (P2, P3, P4 and P6) was previously published as part of our earlier demonstration of parametric responses to rotation symmetry in wallpaper groups\cite{RN1725}. We replicate that result using our Bayesian approach, and find the same parametric effect in the psychophysical data (Supplementary Figure 4.1 - NEEDS TO BE LABELED IN THE SM). We also conducted analyses looking for effects of index and normality in our two datasets, and found that ... Finally, we conducted a correlation analysis comparing SSVEP and behavioral data, and found a small ($R^2 = 0.44$) but above-zero correlation, as indicated by our confidence intervals. There are several factors that might explain the relatively weak correlation, most prominently the fact that the same individuals did not participate in each of the two experiments. Nevertheless, we find the relationship between the two datasets interesting, because it suggests that our psychophysical and SSVEP measurements are tapping into the same underlying mechanisms.

\begin{figure}[tbhp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean RMS SSVEP response. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that xx/64 subgroup relationships have $p(\Delta|data)>0.9$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[tbhp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean display duration threshold. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that xx/64 subgroup relationships have $p(\Delta|data)>0.9$.}
\label{fig:durations_comp}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \cite{RN1170} and rotation \cite{RN1725}, the visual system explicitly represents hierarchical structure of the 17 wallpaper groups, and thus the compositions of all four of the fundamental symmetry transformations (rotation, reflection, translation, glide reflection) which comprise regular textures. The RMS measure of SSVEP amplitude, preserves the complex hierarchy of subgroup relationships among the wallpaper groups \cite{RN1711}. This remarkable consistency was specific to the odd harmonics of the stimulus frequency, that capture the symmetry-specific response \cite{RN1725} and to electrodes in an ROI over occipital cortex. When the same analysis was done using the odd harmonics from a parietal cortex ROI (Supplementary Figure 3.4) or using the even harmonics of the occipital cortex ROI (Supplementary Figure 3.4), the data was much less consistent with the subgroup relationships (yellow and green lines, Figure 4).

The current data provide a complete description of the visual system’s response to symmetries in the 2-D plane. Our design does not allow us to independently measure the response to P1, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 1), we can conclude that the relationships between P1 and all other groups, where P1 is the subgroup, are also preserved by the visual system. The subgroup relationships are not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Furthermore, behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \cite{RN1253}, they are generally error-prone when asked to assign exemplar images to the appropriate group \cite{RN172}. The correspondence between responses in the visual system and group theory that we demonstrate here, may reflect a form of implicit learning that depends on the structure of the natural world. The environment is itself constrained by physical forces underlying pattern formation and these forces are subject to multiple symmetry constraints \cite{RN1634}. The ordered structure of responses to wallpaper groups could be driven by a central tenet of neural coding, that of efficiency. If coding is to be efficient, neural resources should be distributed in such a way that the structure of the environment is captured with minimum redundancy considering the visual geometric optics, the capabilities of the subsequent neural coding stages and the behavioral goals of the organism \cite{RN1758, RN1760, RN1757, RN1756}. Early work within the efficient coding framework suggested that natural images had a $1/f$ spectrum and that the corresponding redundancy between pixels in natural images could be coded efficiently with a sparse set of oriented filter responses, such as those present in the early visual pathway \cite{RN1740, RN1446}. Our results suggest that the principle of efficient coding extends to a much higher level of structural redundancy – that of symmetries in visual images. The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \cite{RN1519}. The correspondence between brain data and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \cite{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \cite{RN1768}. Field biologist have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \cite{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered super-textures, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \cite{RN1764}. Wallpaper groups are visually compelling and have been widely used in human artistic expression going back to the Neolithic age \cite{RN1949}. If wallpapers are super-textures, their prevalence may be a direct consequence of the strategy the human visual system uses for encoding visual textures.

\begin{figure}[tbhp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relations that satisfy $p(\Delta >0 | data) > x$. We can see that if we take $x = 0.95$ as our threshold, the subgroup relations are preserved in 56/63 = 89\% and 49/64 = 78\% of the comparisons for the primary EEG and display durations repectively. This compares to the 32/64= 50\% and 22/64 = 35\% for the control EEG conditions.}
\label{fig:durations_rotations}
\end{figure}

@@

\section*{snippets}
Specifically, the amplitudes of symmetry-specific responses in individual participants ($n=25$) preserve these relationships at an above-chance level in 88.3\% (53 out of 60) of cases.

we used a steady-state design, in which exemplar images belonging to 16 of the 17 wallpaper groups alternated with phase-scrambled images of the same group. 
Exemplars from each of the 16 groups alternated at 0.83 Hz with their corresponding set of P1 exemplars, that were matched in terms of their Fourier power spectrum.

Thus, the magnitude of the odd harmonic response components can be used as a distance metric for each group, with distance being measured relative to the simplest group, P1.

>>>>

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission
% \usepackage{graphicx}

@@

\maketitle
 \thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

@@

\begin{figure}[ptb]
\centering
\includegraphics[width=1\linewidth]{../figures/figure1.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a P1 exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:example_stimuli}
\end{figure}

@@

We now wanted to quantify the degree to which our two measurements were consistent with the subgroup relationships. We hypothesized that more complex groups would (a) produce symmetric-specific SSVEPs with higher amplitudes and (b) shorter display duration thresholds. \textcolor{red}{PERHAPS WE NEED TO JUSTIFY THESE ASSUMPTIONS?} We tested each of these hypotheses using the same approach. We first fitted a Bayesian model with wallpaper group as a factor and participant as a random effect. We fit the model separately for SSVEP RMS and psychophysical data, and then computed posterior distributions for the difference between supergroup and subgroup. These difference distributions could allowed us to compute the conditional probability that the supergroup would produce (a) larger RMS and (b) a shorter threshold durations, when compared to the subgroup. The posterior distributions are shown in Figure 2 for the SSVEP data, and in Figure 3 for the psychophysical data, which distributions color-coded according to conditional probability. For both data sets our hypothesis is confirmed: For the overwhelming majority of the 64 subgroup relationships, supergroups are more likely to produce larger symmetry specific SSVEPs and shorter threshold durations, and in most cases the conditional probability of this happening is extremely high.

@@

SSVEP data from four of the wallpaper groups (P2, P3, P4 and P6) was previously published as part of our earlier demonstration of parametric responses to rotation symmetry in wallpaper groups\cite{RN1725}. We replicate that result using our Bayesian approach, and find the same parametric effect in the psychophysical data (Supplementary Figure 4.1). We also conducted an analysis \textcolor{red}{DOES THIS NEED TO BE SPELLED OUT MORE} looking for effects of index and normality in our two datasets, and found that subgroup relationships with higher indices tended to produce stronger effects for both SSVEP RMS and symmetry detection thresholds, as shown by the fact that a leave-one-out test \textcolor{red}{... MORE DETAILS NEEDED HERE}. The effect of index did not (\textcolor{red}{???} reach significance.

Finally, we conducted a correlation analysis comparing SSVEP and psychophysical data, and found a small ($R^2 = 0.44$) but above-zero correlation, as indicated by our confidence intervals. There are several factors that might explain the relatively weak correlation, most prominently the fact that the same individuals did not participate in each of the two experiments. Nevertheless, we find the relationship between the two datasets interesting, because it suggests that our psychophysical and SSVEP measurements are tapping into the same underlying mechanisms.

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean RMS SSVEP response. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that xx/64 subgroup relationships have $p(\Delta|data)>0.9$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[tbh]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean display duration threshold. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that xx/64 subgroup relationships have $p(\Delta|data)>0.9$.}
\label{fig:durations_comp}
\end{figure}

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \cite{RN1170} and rotation \cite{RN1725}, the visual system explicitly represents hierarchical structure of the 17 wallpaper groups, and thus the compositions of all four of the fundamental symmetry transformations (rotation, reflection, translation, glide reflection) which comprise regular textures. The SSVEP amplitude and the symmetry detection threshold both preserve the complex hierarchy of subgroup relationships among the wallpaper groups \cite{RN1711}. For the SSVEP, this remarkable consistency was specific to the odd harmonics of the stimulus frequency, that capture the symmetry-specific response \cite{RN1725} and to electrodes in an ROI over occipital cortex. When the same analysis was done using the odd harmonics from electrodes over parietal cortex (Supplementary Figure 3.3) or even harmonics from electrodes over occipital cortex (Supplementary Figure 3.4), the data was much less consistent with the subgroup relationships (yellow and green lines, Figure 4).

The current data provide a complete description of the visual system’s response to symmetries in the 2-D plane. Our design precludes us from independently measure the response to P1, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 1), we can conclude that the relationships between P1 and all other groups, where P1 is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Furthermore, previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \cite{RN1253}, they are generally error-prone when asked to assign exemplar images to the appropriate group \cite{RN172} \textcolor{red}{ALASDAIR PERHAPS YOU WANT TO SAY MORE OR DIFFERENT HERE}.

The correspondence between responses in the visual system and group theory that we demonstrate here, may reflect a form of implicit learning that depends on the structure of the natural world. The environment is itself constrained by physical forces underlying pattern formation and these forces are subject to multiple symmetry constraints \cite{RN1634}. The ordered structure of responses to wallpaper groups could be driven by a central tenet of neural coding, that of efficiency. If coding is to be efficient, neural resources should be distributed in such a way that the structure of the environment is captured with minimum redundancy considering the visual geometric optics, the capabilities of the subsequent neural coding stages and the behavioral goals of the organism \cite{RN1758, RN1760, RN1757, RN1756}. Early work within the efficient coding framework suggested that natural images had a $1/f$ spectrum and that the corresponding redundancy between pixels in natural images could be coded efficiently with a sparse set of oriented filter responses, such as those present in the early visual pathway \cite{RN1740, RN1446}. Our results suggest that the principle of efficient coding extends to a much higher level of structural redundancy – that of symmetries in visual images.

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \cite{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \cite{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \cite{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \cite{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered super-textures, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \cite{RN1764}. Wallpaper groups are visually compelling and have been widely used in human artistic expression going back to the Neolithic age \cite{RN1949}. If wallpapers are in fact super-textures, it would imply that this prevalence may be a direct consequence of the strategy the human visual system uses for encoding textures.

\begin{figure}[tbp]
\centering
\includegraphics[width=1\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relations that satisfy $p(\Delta >0 | data) > x$. We can see that if we take $x = 0.95$ as our threshold, the subgroup relations are preserved in 56/63 = 89\% and 49/64 = 78\% of the comparisons for the primary EEG and display durations repectively. This compares to the 32/64= 50\% and 22/64 = 35\% for the control EEG conditions.}
\label{fig:durations_rotations}
\end{figure}

@@

%\section*{snippets}
%Specifically, the amplitudes of symmetry-specific responses in individual participants ($n=25$) preserve these relationships at an above-chance level in 88.3\% (53 out of 60) of cases.

%we used a steady-state design, in which exemplar images belonging to 16 of the 17 wallpaper groups alternated with phase-scrambled images of the same group. 
%Exemplars from each of the 16 groups alternated at 0.83 Hz with their corresponding set of P1 exemplars, that were matched in terms of their Fourier power spectrum.

%Thus, the magnitude of the odd harmonic response components can be used as a distance metric for each group, with distance being measured relative to the simplest group, P1.
